---
title: "Lab 2. DataFrame"
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
    df_print: paged
date: "2023-08-02"
editor_options:
  markdown:
    wrap: 79
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
import::from(magrittr, "%>%", "%$%", .into = "operators")
```

## Setup

```{r setup, eval = FALSE}
devtools::install_github("JoeyBernhardt/singer")
```

## R Markdown

```{r}
# library(tidyverse)
library(singer)
```

## Including Plots

We can obtain a quick overview of what dataset are now available via `singer`
using the `data()` function.

```{r}
data(package = "singer")
```

```{r}
data(singer_locations)
```

## Introduction

```{r}
singer_locations %>% nrow()
```

We can use `head()`, `tail()` or the beautiful `glimpse()` to peek at the
dataset.

```{r}
singer_locations %>% head()
```

```{r}
singer_locations %>% dplyr::glimpse()
```

## Selecting

```{r}
singer_locations %>%
  dplyr::select(title) %>% # let's see only the title column
  head() # and just a few rows
```

Moreover, `tidyverse` (specifically, `dplyr`) offers you helper functions to
select some column, for example:

```{r}
singer_locations %>%
  dplyr::select(dplyr::ends_with("id")) %>%
  head()
```

And we can also decide which columns NOT to have:

```{r}
singer_locations %>%
  dplyr::select(-dplyr::ends_with("id")) %>%
  head()
```

## Plot, listen, 3D print

```{r}
singer_locations %>% skimr::skim()
```

```{r}
singer_locations %>%
  skimr::skim() %>%
  knitr::kable()
```
